Chaotic dispersal of tidal debris

### 2015-07-30 ms.tex
kvj final comments and xsede ack

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the axisymmetric potential. It is well known that the number of degrees of freedom of a potential plays a critical role in determining the orbit structure of the potential; Hamiltonians with more than two degrees of freedom generically contain significant chaotic regions. \citet{pearson15} tested the stochasticity of the orbit of the progenitor that produced `fanned' debris by computing the Lyapunov exponent along this orbit but found that it is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown previously that along some strongly chaotic orbits and in live cosmological haloes, tidal streams do form large, diffuse `fans' of debris \citep[e.g.,][]{fardal14, ngan15}, however it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity. The result from \citet{pearson15} suggests that even weak chaos (as measured by the Lyapunov exponent) may affect the density evolution and therefore observability of tidal streams. Understanding why this occurs and developing a measure to quantify the importance of this enhanced density evolution is a promising new direction to be explored further.

In this \emph{Article}, we study the effect of chaotic diffusion of the fundamental frequencies of individual orbits on the density evolution of tidal debris. We choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics for grids of constant-energy orbits, and compare these results to measures of the density evolution of finite-volume ensembles of orbits (meant to mimic tidal debris). We find that even when the chaotic timescale is predicted to be long for a given orbit, chaos may manifest over much shorter times in small orbit ensembles through the chaotic diffusion of the individual orbits. For a chaotic orbit, the frequency spectrum of the orbit evolves with time: For a small ensemble, the spread in frequencies is therefore time-dependent, which could enhance phase-mixing. This idea supports a reevaluation of the importance of chaos in galactic haloes and implies that the amount of chaos in a given potential may have significant consequences for the observability and survivability of thin, cold tidal streams. We propose a new measure for determining whether this small-scale frequency diffusion will be important for the morphological evolution of tidal debris with a given spread in frequencies or orbital properties over finite and dynamically relevant times.

This paper is organized as follows. We review relevant nonlinear dynamics in Section~\ref{sec:nldreview}. In Section~\ref{sec:methods}, we describe our choice of potential, method for numerical orbit integration, and introduce the chaos indicators used in this work. Our results are split into four subsections: in Section~\ref{sec:results1} we present iso-energy grids of orbits and discuss the orbit classes and chaotic timescales present in our potential; in Section~\ref{sec:results2} we study the density evolution of small ensembles of orbits around each orbit of the previous section; in Section~\ref{sec:results3} we describe the behavior of chaotic diffusion and use this to explain how chaos is relevant for tidal streams over short times. We discuss the implications of our results in Section~\ref{sec:discussion}, and conclude in Section~\ref{sec:conclusions}.

@@

We integrate all orbits in the grid for 10000 orbital periods and use the method described in Section~\ref{sec:lyap} to compute the Lyapunov exponents. Figure~\ref{fig:lyapmap} again shows the grid of initial conditions, but now the color corresponds to the logarithm of the Lyapunov time. The darker pixels have shorter Lyapunov times and are more chaotic. Because of the fixed the integration time, the MLE cannot detect weak chaos and the majority of orbits appear to be regular because they have exceedingly long Lyapunov times (all white points have $(t_\lambda/\periods) \gtrsim 1000$).

@@

The Lyapunov and frequency diffusion times measure the timescales over which chaos is relevant for a given orbit---that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit, or of how long it takes for the frequencies of a single orbit to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy or frequency-space dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^5$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\gtrsim$10\%. In this section, we ask whether  the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent.

It is computationally intractable to run full N-body simulations for the large grid of orbital initial conditions of the previous section and we therefore take a simplified approach for studying how finite-volume debris spreads along each of these orbits. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions---the `parent' orbit---we integrate for 128 orbital periods, find the phase-space position of the minimum pericenter over this time, initialize a small ensemble of test particle orbits around this position, and integrate the orbits of all test particles for some integration time. We are interested in the degree to which chaos enhances the mixing rate of orbit ensembles and we therefore want to isolate out the effect of ordinary phase-mixing along regular orbits; all ensembles should start with the same initial spread in fundamental frequencies. We set the physical scale of the ensemble by the tidal radius in position and the velocity scale in velocity; the initial spread in fundamental frequencies will scale as $(m/M)^{1/3}$ like the tidal radius and velocity scale \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x},\bs{v})$ are a set of initial conditions at pericenter for a parent orbit, then $\delta x_{k,i}$ and $\delta v_{k,i}$ are the $k$th components ($k=1,2,3$) of the position and velocity deviation vectors for the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta x_{k,i} &\sim \mathcal{N}(0, r_{\rm tide}/\sqrt{3})\\
	\delta v_{k,i} &\sim \mathcal{N}(0, \sigma_v/\sqrt{3})\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{M(<\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{M(<\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M(<r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the `progenitor,' and $\|\cdot \|$ is the Euclidean norm. We take $m=10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass of the host potential to estimate the above debris scales. % we have verified this using regular orbits in a logarithmic potential...

We start by considering four particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular, near-resonant orbit (A), a regular, non-resonant orbit (B), a weakly chaotic orbit (C), and a strongly chaotic orbit (D). The orbits were chosen to be close on the orbit grid so that their orbital properties (e.g., apocenter, pericenter) are similar, but have different frequency diffusion times; all four orbits are long-axis tube orbits. Figure~\ref{fig:orbits} shows the orbits in projection over a long integration period (1024 orbital periods). The initial conditions and chaos diagnostics for each orbit are listed in Table~\ref{tbl:orbit-info}. Figure~\ref{fig:ensembles} shows final positions of test-particle ensembles initialized around the four orbits described above and evolved for 64 orbital periods. A thin stream forms on the near-resonant orbit (left column), a thin---but more two-dimensional---stream forms on the non-resonant orbit (middle-left), a more diffuse stream forms on the weakly chaotic orbit with an interesting morphology (middle-right), and a `fanned' stream on the strongly chaotic orbit (right). Given the long Lyapunov and frequency diffusion times of the weakly chaotic orbit ($t_\lambda/\periods > 900$, $t_\Omega/\periods \approx 2\times10^5$), it is surprising that the density evolution of the ensemble on this orbit appears to be more diffuse than the regular orbit stream.

@@

\begin{table*}[ht]
\begin{center}
	\begin{tabular}{c | c c c c c c | c c }
		{\bf ID} & $\bs{x/r_s}$ & $\bs{y/r_s}$ & $\bs{z/r_s}$ & $\bs{v_x/v_c}$ & $\bs{v_y/v_c}$ & $\bs{v_z/v_c}$ & $\bs{t_\lambda/\periods}$ & $\bs{t_\Omega/\periods}$ \\\toprule
A & 0.85 & 0 & 1.303 & 0 & 0.721 & 0 & $>700$ & $>10^7$\\
\midrule
B & 0.85 & 0 & 1.152 & 0 & 0.849 & 0 & $>700$ & $>10^7$\\
\midrule
C & 0.85 & 0 & 1.27 & 0 & 0.752 & 0 & $>700$ & $\approx 3\times10^5$\\
\midrule
D & 0.85 & 0 & 1.434 & 0 & 0.597 & 0 & 8.14 & $\approx 2.5\times10^4$\\
		\bottomrule
		\end{tabular}
	\caption{Orbital initial conditions from the $xz$ grid for orbits with a range of chaotic timescales---A is a regular, near-resonant orbit, B a regular, non-resonant orbit, C a weakly chaotic orbit, and D a strongly chaotic orbit. Positions ($x$, $y$, $z$) are given in units of scale radii, $r_s$, and velocities ($v_x$, $v_y$, $v_z$) in units of scale velocity, $v_c$. Chaotic timescales are expressed ($t_\lambda$, $t_\Omega$) in number of orbital periods. Recall that the frequency diffusion time, $t_\Omega$, is the time over which we expect order-unity changes in the fundamental frequencies, hence why the timescales appear quite long. \label{tbl:orbit-info}}
\end{center}
\end{table*}

@@

Motivated by the noticeable discrepancy between the final mean density between the two regular (A,B) and the weakly chaotic (C) orbit ensembles---even though the chaotic timescale of the weakly chaotic ensemble parent orbit is several times the integration period used above---we compute the final mean density for orbit ensembles generated around each orbit in the grid described in Section~\ref{sec:results1}. For all ensembles, we integrate the orbits for 64 (parent) orbital periods and compute the initial and final values of the mean, configuration-space density. Figure~\ref{fig:ensemblemap-meandensity} again shows the grid of initial conditions (Section~\ref{sec:results1}, but now the greyscale indicates the ratio of the final mean density to the initial density. Interestingly, much of the structure that is visible in the upper-half of the frequency diffusion time map (Figure~\ref{fig:freqdiff}) is again visible in this map of the density evolution of orbit ensembles. Many stable resonances appear as lighter, curved features in the upper portion of this grid where the density remains systematically higher, and regions of strong resonance overlap lead to weak chaos and systematically lower density. However, it is surprising that these features are present: The chaotic timescales predicted from both the Lyapunov and frequency diffusion times were typically 100 to 1000s of orbital periods for many of the unstable features around resonances. The timescales in the lower portion of the grid (where the orbits are primarily stable, short-axis tube orbits) are simply too long to detect density differences over 64 orbital periods even from weak chaos in this particular choice of potential.

@@

As a demonstration of the small-scale frequency modulation, we consider again the four orbits of Section~\ref{sec:results2} (initial conditions are listed in Table~\ref{tbl:orbit-info}). To resolve the short-time behavior of the frequency diffusion (corresponding to motion across or around resonance layers) we compute the frequencies in a series of rolling windows along each orbit. We use a window with a width equal to 64 orbital periods and shift the window by an orbital period between each calculation of the most significant frequencies. Figure~\ref{fig:four-orbits-freqs} shows plots of the frequencies of each orbit computed in each window of time---plotted in projection are the percent deviations of the frequency from the initial value. The difference between the first window (green +) and the last window (red x) is 128 orbital periods. For the two regular orbits, (A) and (B), the frequencies vary only from numerical issues when recovering the frequencies ($\ll$1\%) and thus the initial and final value markers overlap. The frequencies of the weakly chaotic orbit (C) evolve quickly but are bounded to a small volume with fractional size $\approx$1\% (presumably by nearby resonant surfaces); this is larger than the frequency spread in globular cluster tidal debris (0.1--0.5\%). The frequencies of the strongly chaotic orbit (D) also evolve quickly but fill a larger volume.

@@

We have shown that the Lyapunov and frequency diffusion times are indicators of chaos and that the frequency diffusion time resolves the detailed resonant structure of gravitational potentials, but the timescales predicted do not capture the importance of chaos for the density evolution of ensembles of orbits meant to mimic tidal debris. We have shown that small-scale but fast chaotic diffusion of orbits can explain this enhanced mixing, and have proposed a new indicator to characterize the importance of this effect over given evolution times. In the sub-sections below, we discuss a few important limitations that remain for exploration in future work.

@@

We have only considered low-mass progenitor systems such as globular clusters because the intrinsic spreads in fundamental frequencies are small (0.1--0.5\%). Small changes to the frequencies of the orbits of tidal debris disrupted from these progenitors due to chaotic processes will therefore cause observable changes to the real-space morphology of the debris. For more massive progenitor systems, the typical size and velocity dispersion of the debris will be larger and thus the debris morphology will be less sensitive to small changes in orbital properties. The mass-scale of the debris that will display enhanced density evolution depends on the magnitude of weak chaos, which depends on the orbit structure of a given potential. In potentials with more significant chaos, debris disrupted from more massive progenitor systems may also display `stream-fanning.'

@@

The Ophiuchus stream \citep{bernard14, sesar15} appears to be a thin, short tidal stream (deprojected length $\approx$1.5--2 kpc) near the Galactic bulge with no apparent progenitor. At Galactocentric $R \approx 1$ kpc, $z \approx 5$ kpc, the orbits of the stream stars likely pass through the MW disk, feel the triaxiality of the Galactic bar \citep[e.g.,][]{wegg13, wegg15}, and have short orbital periods (relative to streams in the halo). It is possible that the observed debris is the last remnants of the recently disrupted progenitor system \citep{sesar15}, however if a significant number of stars were disrupted on previous pericentric passages, this older debris may be `fanned' and still near the observed portion of the stream. The fanned debris would be much lower surface brightness and thus much more difficult to detect. The detection of this low-surface-brightness component would open up the possibility that enhanced density evolution due to chaos is a dynamically relevant process for thin streams in real galaxies, though \cite{carlberg15} has shown that it is also possible also to get short, high density segments of streams from streams formed on eccentric orbits.

@@

APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ DGE 11-44155. This work was partially supported by the National Science Foundation under Grant No. AST-1312196. MV is supported in part by the University of MichiganÕs Elizabeth Crosby Fund 
and the Office of the VP for Research and NASA ATP grant NNX15AK79G. AHWK would like to acknowledge support from NASA through Hubble Fellowship grant HST-HF-51323.01-A awarded by the Space Telescope Science Institute, which is operated by the Association of Universities for Research in Astronomy, Inc., for NASA, under contract NAS 5-26555.

This research made use of \project{Astropy}, a community-developed core \project{Python} package for Astronomy \citep{astropy13}. This work additionally relied on Columbia University's \emph{Hotfoot} and \emph{Yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance.

@@

% Check frequency analysis precision
\clearpage
\begin{figure}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/freq-evolution.pdf}
\caption{Evolution of the fundamental frequencies computed over a total of 256 orbital periods for the four orbits, A, B, C, D. Plotted are the per cent deviations of the frequencies computed in window relative to the value in the initial window---that is, if $j$ is the index of a given time window, $\delta \Omega_{1,j} = (\Omega_{1,j} - \Omega_{1,0})/\Omega_{1,0}$. The initial value is shown as a green plus sign and the final value is shown as a red x. The frequencies are computed with a window width of $\approx$128 orbital periods, and the window is shifted by one orbital period between each computation (each grey point represents the fundamental frequencies computed in a single window). For the regular orbits (A,B), the fractional variation is around $10^{-6}$ and thus all points overlap on this scale. The weakly chaotic orbit (C) displays frequency variations comparable to the spread in frequencies in globular-cluster-like tidal debris. The frequency variations of the strongly chaotic orbit (D) have a larger characteristic spread.} 
\label{fig:four-orbits-freqs}
\end{center}
\end{figure}

>>>>

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the oblate potential. It is well known that the number of degrees of freedom of a potential plays a critical role in determining the orbit structure of the potential; Hamiltonians with more than two degrees of freedom generically contain significant chaotic regions. \citet{pearson15} tested the stochasticity of the orbit of the progenitor that produced `fanned' debris by computing the Lyapunov exponent along this orbit but found that it is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown previously that along some strongly chaotic orbits and in live cosmological haloes, tidal streams do form large, diffuse `fans' of debris \citep[e.g.,][]{fardal14, ngan15}, however it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity. The result from \citet{pearson15} suggests that even weak chaos (as measured by the Lyapunov exponent) may affect the density evolution and therefore observability of tidal streams. Understanding why this occurs and developing a measure to quantify the importance of this enhanced density evolution is a promising new direction to be explored further.

In this \emph{Article}, we study the effect of chaotic diffusion of the fundamental frequencies of individual orbits on the density evolution of tidal debris. We choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics for grids of constant-energy orbits, and compare these results to measures of the density evolution of finite-volume ensembles of orbits (meant to mimic tidal debris). We find that even when the chaotic timescale is predicted to be long for a given orbit, chaos may manifest over much shorter times in small orbit ensembles through the chaotic diffusion of the individual orbits. For a chaotic orbit, the frequency spectrum of the orbit evolves with time: for a small ensemble, the spread in frequencies is therefore time-dependent, which could enhance phase-mixing. This idea supports a reevaluation of the importance of chaos in galactic haloes and implies that the amount of chaos in a given potential may have significant consequences for the observability and survivability of thin, cold tidal streams.

This paper is organized as follows. We review relevant nonlinear dynamics in Section~\ref{sec:nldreview}. In Section~\ref{sec:methods}, we describe our choice of potential, method for numerical orbit integration, and introduce the chaos indicators used in this work. Our results are split into three subsections: in Section~\ref{sec:results1} we present iso-energy grids of orbits and discuss the orbit classes and chaotic timescales present in our potential; in Section~\ref{sec:results2} we study the density evolution of small ensembles of orbits around each orbit of the previous section; in Section~\ref{sec:results3} we describe the behavior of chaotic diffusion and use this to explain how chaos is relevant for tidal streams over short times. We discuss the implications of our results in Section~\ref{sec:discussion}, and conclude in Section~\ref{sec:conclusions}.

@@

We integrate all orbits in the grid for 10000 orbital periods and use the method described in Section~\ref{sec:lyap} to compute the MLEs. Figure~\ref{fig:lyapmap} again shows the grid of initial conditions, but now the color corresponds to the logarithm of the inverse of the MLE (the Lyapunov time). The darker pixels have shorter Lyapunov times and are more chaotic. Because of the fixed the integration time, the MLE cannot detect weak chaos and the majority of orbits appear to be regular because they have exceedingly long Lyapunov times (all white points have $(t_\lambda/\periods) \gtrsim 1000$).

@@

The Lyapunov and frequency diffusion times measure the timescales over which chaos is relevant for a given orbit---that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit, or of how long it takes for the frequencies of a single orbit to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy or frequency-space dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^6$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\sim$10\%. In this section, we ask whether  the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent.

It is computationally intractable to run full N-body simulations for the large grid of orbital initial conditions of the previous section and we therefore take a simplified approach for studying how finite-volume debris spreads along each of these orbits. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions---the `parent' orbit---we integrate for 128 orbital periods, find the phase-space position of the minimum pericenter over this time, initialize a small ensemble of test particle orbits around this position, and integrate the orbits of all test particles for some integration time. We are interested in the degree to which chaos enhances the mixing rate of orbit ensembles and we therefore want to isolate out the effect of ordinary phase-mixing along regular orbits; all ensembles should start with the same initial spread in fundamental frequencies. We set the physical scale of the ensemble by the tidal radius in position and the velocity scale in velocity; the initial spread in fundamental frequencies will scale as $(m/M)^{1/3}$ like the tidal radius and velocity scale \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are a set of initial conditions at pericenter for a parent orbit, then $\delta x_{k,i}$ and $\delta v_{k,i}$ are the $k$th components ($k=1,2,3$) of the position and velocity deviation vectors for the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta x_{k,i} &\sim \mathcal{N}(0, r_{\rm tide}/\sqrt{3})\\
	\delta v_{k,i} &\sim \mathcal{N}(0, \sigma_v/\sqrt{3})\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{M(<\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{M(<\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M(<r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the `progenitor,' and $\|\cdot \|$ is the Euclidean norm. We take $m=10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass of the host potential to estimate the above debris scales. % we have verified this using regular orbits in a logarithmic potential...

We start by considering four particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular, near-resonant orbit (A), a regular, non-resonant orbit (B), a weakly chaotic orbit (C), and a strongly chaotic orbit (D). The orbits were chosen to be close on the orbit grid so that their orbital properties (e.g., apocenter, pericenter) are similar, but have different frequency diffusion times; all four orbits are long-axis tube orbits. Figure~\ref{fig:orbits} shows the orbits in projection over a long integration period (1024 orbital periods). The initial conditions and chaos diagnostics for each orbit are listed in Table~\ref{tbl:orbit-info}. Figure~\ref{fig:ensembles} shows final positions of test-particle ensembles initialized around the four orbits described above and evolved for 64 orbital periods. A thin stream forms on the near-resonant orbit (left column), a thin---but more two-dimensional---stream forms on the non-resonant orbit (middle-left), a more diffuse stream forms on the weakly chaotic orbit with a slightly `fanned' morphology (middle-right), and a two-dimensional, `fanned' stream on the strongly chaotic orbit (right). Given the long Lyapunov and frequency diffusion times of the weakly chaotic orbit ($t_\lambda/\periods > 900$, $t_\Omega/\periods \approx 2\times10^5$), it is surprising that the density evolution of the ensemble on this orbit appears to be more diffuse than the regular orbit stream.

@@

\begin{table*}[ht]
\begin{center}
	\begin{tabular}{c | c c c c c c | c c}
		{\bf ID} & $\bs{x/r_s}$ & $\bs{y/r_s}$ & $\bs{z/r_s}$ & $\bs{v_x/v_c}$ & $\bs{v_y/v_c}$ & $\bs{v_z/v_c}$ & $\bs{t_\lambda/\periods}$ & $\bs{t_\Omega/\periods}$ \\\toprule
A & 0.85 & 0 & 1.303 & 0 & 0.721 & 0 & $>700$ & $>10^7$ \\
\midrule
B & 0.85 & 0 & 1.152 & 0 & 0.849 & 0 & $>700$ & $>10^7$\\
\midrule
C & 0.85 & 0 & 1.27 & 0 & 0.752 & 0 & $>700$ & $\approx 3\times10^5$\\
\midrule
D & 0.85 & 0 & 1.434 & 0 & 0.597 & 0 & 8.14 & $\approx 2.5\times10^4$\\
		\bottomrule
		\end{tabular}
	\caption{Orbital initial conditions from the $xz$ grid for orbits with a range of chaotic timescales---A is a regular, near-resonant orbit, B a regular, non-resonant orbit, C a weakly chaotic orbit, and D a strongly chaotic orbit. Positions ($x$, $y$, $z$) are given in units of scale radii, $r_s$, and velocities ($v_x$, $v_y$, $v_z$) in units of scale velocity, $v_c$. Chaotic timescales are expressed ($t_\lambda$, $t_\Omega$) in number of orbital periods. Recall that the frequency diffusion time, $t_\Omega$, is the time over which we expect order-unity changes in the fundamental frequencies, hence why the timescales appear quite long. \label{tbl:orbit-info}}
\end{center}
\end{table*}

@@

Motivated by the noticeable discrepancy between the final mean density between the two regular (A,B) and the weakly chaotic (C) orbit ensembles---even though the chaotic timescale of the weakly chaotic ensemble parent orbit is several times the integration period used above---we compute the final mean density for orbit ensembles generated around each orbit in the grid described in Section~\ref{sec:results1}. For all ensembles, we integrate the orbits for 64 (parent) orbital periods and compute the initial and final values of the mean, configuration-space density. Figure~\ref{fig:ensemblemap-meandensity} again shows the grid of initial conditions (Section~\ref{sec:results1}, but now the greyscale indicates the ratio of the final mean density to the initial density. Interestingly, much of the structure that is visible in the upper-half of the frequency diffusion time map (Figure~\ref{fig:freqdiff}) is again visible in this map of the density evolution of orbit ensembles. Many features appear as lighter, curved features in the upper portion of this grid where the density remains systematically higher---these are the stable resonances of the potential. Darker regions have systematically lower densities and correspond to regions of resonance overlap where weakly chaotic orbits are found.

However, it is surprising that these features are present: The chaotic timescales predicted from both the Lyapunov and frequency diffusion times were typically 100 to 1000s of orbital periods for many of the unstable features around resonances. The timescales in the lower portion of the grid (where the orbits are primarily stable, short-axis tube orbits) are simply too long to detect density differences over 64 orbital periods even from weak chaos in this particular choice of potential.

@@

As a demonstration of the small-scale frequency modulation, we consider again the four orbits of Section~\ref{sec:results2} (initial conditions are listed in Table~\ref{tbl:orbit-info}). To resolve the short-time behavior of the frequency diffusion (corresponding to motion across or around resonance layers) we compute the frequencies in a series of rolling windows along each orbit. We use a window with a width equal to 64 orbital periods and shift the window by an orbital period between each calculation of the most significant frequencies. Figure~\ref{fig:four-orbits-freqs} shows plots of the frequencies of each orbit computed in each window of time---plotted in projection are the percent deviations of the frequency from the initial value. The difference between the first window (blue +) and the last window (red x) is 128 orbital periods. For the two regular orbits, (A) and (B), the frequencies vary only from numerical issues when recovering the frequencies ($\ll$1\%) and thus the initial and final value markers overlap. The frequencies of the weakly chaotic orbit (C) evolve quickly but are bounded to a small volume with fractional size $\approx$1\% (presumably by nearby resonant surfaces); this is larger than the frequency spread in globular cluster tidal debris (0.1--0.5\%). The frequencies of the strongly chaotic orbit (D) also evolve quickly but fill a larger volume.

@@

We have shown that the Lyapunov and frequency diffusion times are indicators of chaos and that the frequency diffusion time resolves the detailed resonant structure of gravitational potentials, but the timescales predicted do not capture the importance of chaos for the density evolution of ensembles of orbits meant to mimic tidal debris. We have shown that small-scale but fast chaotic diffusion of orbits can explain this enhanced mixing. In the sub-sections below, we discuss a few important limitations that remain for exploration in future work.

@@

We have only considered low-mass progenitor systems such as globular clusters because the intrinsic spreads in fundamental frequencies are small (0.1--0.5\%). Small changes to the frequencies of the orbits of tidal debris disrupted from these progenitors due to chaotic processes will therefore cause observable changes to the real-space morphology of the debris. For more massive progenitor systems, the typical size and velocity dispersion of the debris will be larger and thus the debris morphology will be less sensitive to small changes in orbital properties. The mass-scale of the debris that will display enhanced density evolution depends on the magnitude of weak chaos, which depends on the orbit structure of a given potential. In potentials with more significant chaos, debris disrupted from more massive progenitor systems may also display `stream-fanning.' However, since the scale of the debris is larger it is more likely that some orbits will exist in chaotic regions.

@@

The Ophiuchus stream \citep{bernard14, sesar15} appears to be a thin, short tidal stream (deprojected length $\approx$1.5--2 kpc) near the Galactic bulge with no apparent progenitor. At Galactocentric $R \approx 1$ kpc, $z \approx 5$ kpc, the orbits of the stream stars likely pass through the MW disk, feel the triaxiality of the Galactic bar \citep[e.g.,][]{wegg13, wegg15}, and have short orbital periods (relative to streams in the halo). It is possible that the observed debris is the last remnants of the recently disrupted progenitor system \citep{sesar15}, however if a significant number of stars were disrupted on previous pericentric passages, this older debris may be `fanned' and still near the observed portion of the stream. The fanned debris would be much lower surface brightness and thus much more difficult to detect. The detection of this low-surface-brightness component would open up the possibility that enhanced density evolution due to chaos is a dynamically relevant process for thin streams in real galaxies, though \cite{carlberg15} has shown that it is also possible also to get short, high density segments of streams from debris formed on eccentric orbits.

@@

APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ DGE 11-44155. KVJ and APW were partially supported by the National Science Foundation under Grant No. AST-1312196 and NASA grant NNX15AK78G. MV is supported in part by the University of MichiganÕs Elizabeth Crosby Fund and the Office of the VP for Research and NASA ATP grant NNX15AK79G. AHWK would like to acknowledge support from NASA through Hubble Fellowship grant HST-HF-51323.01-A awarded by the Space Telescope Science Institute, which is operated by the Association of Universities for Research in Astronomy, Inc., for NASA, under contract NAS 5-26555.

This research made use of \project{Astropy}, a community-developed core \project{Python} package for Astronomy \citep{astropy13}. This work additionally relied on Columbia University's \emph{Hotfoot} and \emph{Yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance. This work used the Extreme Science and Engineering Discovery Environment (XSEDE), which is supported by National Science Foundation grant number ACI-1053575 \citep{xsede}.

@@

% Check frequency analysis precision
\clearpage
\begin{figure}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/freq-evolution.pdf}
\caption{Evolution of the fundamental frequencies computed over a total of 256 orbital periods for the four orbits, A, B, C, D. Plotted are the per cent deviations of the frequencies computed in window relative to the value in the initial window---that is, if $j$ is the index of a given time window, $\delta \Omega_{1,j} = (\Omega_{1,j} - \Omega_{1,0})/\Omega_{1,0}$. The initial value is shown as a blue plus sign and the final value is shown as a red x. The frequencies are computed with a window width of $\approx$128 orbital periods, and the window is shifted by one orbital period between each computation (each grey point represents the fundamental frequencies computed in a single window). For the regular orbits (A,B), the fractional variation is around $10^{-6}$ and thus all points overlap on this scale. The weakly chaotic orbit (C) displays frequency variations comparable to the spread in frequencies in globular-cluster-like tidal debris. The frequency variations of the strongly chaotic orbit (D) have a larger characteristic spread.} 
\label{fig:four-orbits-freqs}
\end{center}
\end{figure}
